A user--friendly tool to evaluate the effectiveness of no--take marine reserves

### 2017-12-03 Manuscript.tex
Incorporated changes from reviewers

Available empirical evidence on marine reserve effectiveness is mixed
{[}7{]}. Some studies have shown that MPAs increase biomass {[}4,8{]},
enhance resilience to climatic impacts {[}9,10{]}, and preserve genetic
diversity {[}11{]}. Compared to partially protected MPAs, marine
reserves have higher levels of biomass, density, richness, and larger
organisms {[}3,12--14{]}. These effects are often measured as biological
changes in the area through time and many lack a control site for
comparison {[}15{]}. This approach does not account for other factors
for which one must control {[}16{]} in order to causally attribute a
biological change to the reserve. While some studies have used control
sites, these analyses do not estimate the effect of the reserve, and
often use a control--impact comparison approach that does not address
temporal variability {[}4,8,17--19{]}. A smaller fraction of studies
have used a before--after--control--impact (\emph{i.e.} BACI) design
comparing reserves to control sites before and after implementation
{[}4,20,21{]}, which allows the use of causal inference techniques that
estimate the effect of the reserve.

But even when proper causal inference can be drawn, results are often
idiosyncratic across reserves. Effects of reserves on ecological and
economic outcomes are highly heterogeneous, and often depend on the
specific ecological, economic, and social context. The purpose of this
paper is to describe a user--friendly tool, called ``MAREA'', to
rigorously systematize the evaluation of marine reserve effectiveness.
The tool is in the form of an open-source application that uses
state--of--the--art methods from program evaluation to compare a reserve
to control sites along a number of ecological, economic, and governance
dimensions.

The challenge of how to standardize marine reserve evaluation is not a
new one. The recent IUCN framework ``How is your MPA doing?''
{[}22,23{]} provides a comprehensive list of biological, socioeconomic,
and governance indicators, and insights into how these indicators may be
measured. But this framework stops short of analysis, so provides a user
with little guidance about establishing causal inference about the
reserve. Recent work by Mascia \emph{et al.} {[}24{]} integrates these
three dimensions via the Social Ecological Systems Framework {[}25,26{]}
and suggests the use of causal inference techniques to provide a measure
of the effect of conservation interventions. However, these two novel
approaches do not provide a user-friendly tool that enables
replicability and scalability of the analysis, particularly when used by
the fishers and decision makers themselves.

@@

This paper presents a framework and user--friendly tool to evaluate
marine reserve effectiveness, which incorporates the biological,
socioeconomic, and governance dimensions of any fishery. We first
provide a list of commonly stated management objectives and match them
to appropriate indicators. We then develop a simple approach to
analyzing these indicators building on causal inference techniques
{[}20{]}, which help us understand the effect of management
interventions {[}24,37{]}. To implement the analytical approach in a
user-friendly format, we introduce the Marine Reserve Evaluation
Application (MAREA), an open source, web--based tool that automates the
framework described in this paper and enables its broader use. Finally,
we present a case study on the evaluation of a marine reserve
established by the fishers of Isla Natividad (Mexico) in 2006, to
demonstrate the potential of MAREA.

@@

Throughout this study, we will refer to the stated goals for which a
marine reserve was designed as ``objectives.'' This work was motivated
by the case of Mexico, where 39 reserves have been implemented over the
past five years to achieve objectives such as increasing productivity in
nearby waters or recover overexploited species; most of these reserves
have never been formally evaluated for effectiveness at meeting those
objectives. Thus, our focus was on identifying common objectives of
marine reserves in Mexico. However, a literature review and discussions
with marine reserve researchers worldwide suggest that the objectives
driving Mexican marine reserve implementation are similar to those in
the rest of the world. Thus, we group these objectives into seven major
categories, which can be applied to marine reserves worldwide. The list
of objectives includes stated objectives in legislation {[}38,39{]} and
official documents such as the Technical Justification Studies
(\emph{Estudios Técnicos Justificativos}), agreements, and decrees
associated to these areas:

@@

Based on these seven objectives, we determined a set of associated
indicators to evaluate reserve effectiveness. These indicators are
specific variables on which data could be collected, and analyzed, to
ultimately determine whether the corresponding objective was causally
being achieved by the marine reserve. The list of indicators was
compiled through a review of scientific literature in which we
identified indicators that were used to measure similar objectives. A
first filter eliminated indicators for which baseline data do not
typically exist in Mexico. The preliminary list of indicators was
reviewed at a workshop with participation of members from Mexican
fishery management agencies and non-government organizations. Later,
these were presented to fishers from the Ensenada Fishing Cooperative
(\emph{S.C.P.P. Ensenada}), in El Rosario, Baja California, who provided
input. Our final list of indicators includes those identified in review
works {[}4,40{]}.

Indicators are divided into three main categories: biological,
socioeconomic, and governance (Table 1). The nine biological indicators
focus on fish and invertebrate communities that are evaluated using
underwater ecological surveys performed inside and outside the reserve
(see Data and Analysis section for specific sampling design and
methodologies). Five socioeconomic indicators reflect the performance of
the fishery in terms of landings, income from landings, and availability
of alternative livelihoods. Fifteen governance indicators describe the
governance structures under which the community operates (\emph{e.g.},
access rights to the fishery, number of fishers, legal recognition of
the reserve). Some indicators require a numerical entry (\emph{e.g.}
Fish biomass) while others are more descriptive (\emph{e.g.} Reasoning
for reserve location). Many of them specifically measure an outcome of
the reserve, though some are designed to further the understanding of
the mechanisms driving a reserve's performance. In that sense, most
biological and socioeconomic indicators are outcome variables. On the
other hand, governance indicators are viewed as possible explanatory
variables of reserve performance. Whenever an indicator is applied to
``Target species'', it means that the indicator can be used for all
species (\emph{e.g.} Fish Biomass) and/or for individual species that
are either the conservation target of the reserve or are of particular
economic or ecological interest (\emph{e.g.} Grouper Biomass). Table 1
presents the proposed indicators, and Table 2 shows how objectives are
matched with biological and socioeconomic indicators.

@@

\caption{\label{tab:unnamed-chunk-2}List of indicators to evaluate the effectiveness of no-take marine reserves.}
\centering
\resizebox{\linewidth}{!}{\begin{tabular}[t]{l|l|l|l}
\hline
Code & Indicator & Data type & Unit\\
\hline
\multicolumn{4}{l}{\textbf{Biological}}\\
\hline
\hspace{1em}B1 & Shannon diversity index & Continuous & \\
\hline
\hspace{1em}B2 & Species richness & Discrete & Number of species/transect\\
\hline
\hspace{1em}B3 & Density of mature organisms & Continuous & Percent points\\
\hline
\hspace{1em}B4 & Density* & Continuous & Organisms/transect\\
\hline
\hspace{1em}B5 & Natural Disturbance & Descriptive & \\
\hline
\hspace{1em}B6 & Mean Trophic Level & Continuous & \\
\hline
\hspace{1em}B7 & Biomass* & Continuous & kg/transect\\
\hline
\multicolumn{4}{l}{\textbf{Socioeconomic}}\\
\hline
\hspace{1em}S1 & Total landings* & Continuous & kg\\
\hline
\hspace{1em}S2 & Income from total landings* & Continuous & \$\\
\hline
\hspace{1em}S3 & Alternative economic opportunities & Ordinal & \\
\hline
\multicolumn{4}{l}{\textbf{Governance}}\\
\hline
\hspace{1em}G1 & Access to the fishery & Categorical & \\
\hline
\hspace{1em}G2 & Number of fishers & Discrete & \\
\hline
\hspace{1em}G3 & Legal recognition of reserve & Binary & \\
\hline
\hspace{1em}G4 & Reserve type & Descriptive & \\
\hline
\hspace{1em}G5 & Illegal harvesting & Ordinal & \\
\hline
\hspace{1em}G6 & Management plan & Binary & \\
\hline
\hspace{1em}G7 & Reserve enforcement & Descriptive & \\
\hline
\hspace{1em}G8 & Size of reserve & Discrete & \\
\hline
\hspace{1em}G9 & Reasoning for reserve location & Descriptive & \\
\hline
\hspace{1em}G10 & Membership to fisher organizations & Binary & \\
\hline
\hspace{1em}G11 & Type of fisheries organizations & Categorical & \\
\hline
\hspace{1em}G12 & Representation & Ordinal & \\
\hline
\hspace{1em}G13 & Internal Regulation & Binary & \\
\hline
\hspace{1em}G14 & Perceived Effectiveness & Categorical & \\
\hline
\hspace{1em}G15 & Social Impact of Reserve & Categorical & \\
\hline
\multicolumn{4}{l}{\textsuperscript{*} The indicator is also applied to target species}\\
\end{tabular}}
\end{table}

@@

\caption{\label{tab:unnamed-chunk-3}Management objectives and respective performance indicators. All governance indicators should allways be used.}
\centering
\resizebox{\linewidth}{!}{\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
\hline
Objective & B1 & B2 & B3 & B4 & B4* & B5 & B6 & B7 & B7* & S1 & S1* & S2 & S2* & S3\\
\hline
Avoid overexploitation &  &  & x &  & x & x &  &  & x & x & x & x & x & x\\
\hline
Conserve species under a special protection &  &  & x &  & x & x &  &  & x & x &  & x &  & x\\
\hline
Maintain biological process &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
\hline
Improve fishery production in nearby waters &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
\hline
Preserve biological diversity and the ecosystem &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
\hline
Recover overexploited species & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
\hline
Recover species of economic interest & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
\hline
\multicolumn{15}{l}{\textsuperscript{*} The indicator is applied to target species}\\
\end{tabular}}
\end{table}

@@

In many coastal marine reserves of Mexico, biological data are collected
via underwater visual censuses as part of a reserve's monitoring
program. Scientific divers (which are often local fishermen with
guidance from Civil Society Organizations; CSOs) record fish and
invertebrate richness and abundances, as well as fish total length along
belt transects. Ecological surveys are typically performed annually in
each reserve and corresponding control site(s), before and after the
implementation of the reserve, providing a sampling design that can be
used to draw causal inference. Control sites are areas where habitat is
similar to that of the reserve, but with presence of fishing activity;
in principle these are areas that are otherwise observationally
identical to the reserve site, but where, for presumably random reasons,
a reserve was not implemented. While transect dimensions (\emph{i.e.}
length and width) and sampling methods might vary from study to study,
the general idea remains the same: richness, abundances, and sizes of
organisms are recorded in a study--specific standardized way. For this
reason, MAREA does not assume specific transect dimensions, and
pertinent indicators are calculated per transect (Table 1).

This sampling design for biological data allows us to use causal
inference techniques {[}20,41{]} to evaluate the effect of the reserve
on biological indicators. The hypothesis that the indicators will
respond to implementation of the reserve is tested by analyzing spatial
and temporal changes in each numeric biological indicator (all but B5)
using generalized linear models {[}20{]}. To account for variations in
the environment and survey conditions, covariates that are gathered
during the underwater ecological surveys are included in the
difference-in-differences model with form:

@@

In this model, \(i\), \(t\), and \(z\) are indices for transect, time,
and zone (control or reserve site), respectively. This model allows us
to estimate the change in an indicator (\(I\)) based on the year
(\(Y\)), a dummy variable that indicates treatment (\(Z\); \emph{i.e.}
control or reserve), an interaction between a dummy variable that
indicates pre-- or post--implementation (\(P\)) and treatment (\(Z\)),
and covariates such as bottom temperature (\(T\); in °C), horizontal
visibility during the survey (\(V\); in m), and depth at which survey
was performed (\(D\); in m). \(\epsilon\) represents the error term
associated to the regression. Here, years are modeled as factors, using
the first year as the reference level. This does not impose a linear
structure in the way an indicator changes through time (\emph{i.e.} the
change in biomass between 2006 and 2007 does not have to be the same as
the change between 2015 and 2016). The treatment and implementation
variables, modeled as dummy variables, are coded as Control = 0 and
Reserve = 1; and Pre--implementation = 0 and Post--implementation = 1,
respectively.

Socioeconomic data are often collected by fishers, natural resource
management agencies, or CSOs by recording landings, income, and
sometimes prices for each species. To control for inflation, income is
adjusted with the country's consumer price index {[}42{]}:

@@

While this model does not formally allow for causal inference, we can
still measure changes in mean landings and income before and after the
implementation of the reserve and provide valuable input. For both
models (eq. 1 and eq. 3), we estimate the model coeficients with
ordinary least squares, and calculate heteroskedastic--robust standard
errors.

While biological and some economic data are regularly collected,
governance data are typically not available nor systematically collected
by the community or other organizations. Therefore, we created a survey
specifically designed to collect information needed for the proposed
indicators (B5, S3, and G1--G15). The survey is included as
supplementary material in English (S1 Appendix) and Spanish (S2
Appendix). To analyze governance information, we developed a framework
based on a literature review of common governance structures and their
relation to effectiveness in managing fisheries or marine reserves (S3
Table). This approach has been proven to successfully evaluate
governance structures {[}43{]}. Unlike with biological and socioeconomic
objectives (see eqs 1 and 3), MAREA does not quantitatively analyze
governance information. Rather, it is presented along with the
biological and socioeconomic indicators to provide managers and users
with a more complete description of the reserve.

@@

We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}44{]} using
the Shiny package {[}45{]}, to build an interactive web application
hosted on an open server; the MAREA app can be accessed at
\href{turfeffect.shinyapps.io/marea/}{turfeffect.shinyapps.io/marea}.
While the original version was developed in Spanish because it was aimed
for Mexico and other Latin-American countries, all of its content can be
translated by a translation widget available within the app.

MAREA is designed as a 6--step process, divided in tabs that appear upon
launching the app. The first tab introduces the app and summarizes the
evaluation process. Then, the user selects management objectives, which
MAREA automatically matches to appropriate indicators, based on Table 2.
Users can also manually modify selected indicators based on their
interests and data availability. The user can then load data on one or
more reserves, using standard *.csv text files; sample datasets are
provided within MAREA. Once data have been loaded, MAREA identifies all
reserves in the data, and lets the user select the reserve to be
evaluated. At this point, the user can also specify the year of
implementation of the reserve, reserve dimensions, and indicate target
species that are of particular management interest. MAREA provides the
user with a section to confirm that all the decisions made leading up to
that point are correct. Once the user has confirmed all input data,
objectives, and other information, MAREA performs the formal program
evaluation analyses discussed above. For a typical data set, the
automated analysis step takes less than one second. Finally, the user is
taken to the results tab where all results are presented in a simple
format. The user can also download a more comprehensive technical report
produced in *.pdf format.

@@

The scorecard is produced with functions from the Shinydashboard package
{[}46{]}. The technical report is produced by a parameterized Rmarkdown
document {[}47{]} processed by the knitr package {[}48{]}. Another
feature of MAREA is that the user can choose to share the data. Once the
technical report is downloaded, the information on the reserve, its
management objectives, and all uploaded data are saved into a central
repository. These data can be accessed at any time by any person
interested in acquiring them at \url{github.com/turfeffect/MAREAdata}.

@@

While MAREA is a general tool that can be easily employed to evaluate
the effectiveness of any marine reserve with the required input data, we
illustrate its use here by applying it to one marine reserve near Isla
Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8
Km off the Pacific Coast of the Baja California Peninsula (Fig. 3),
where fishers operate under a fishing cooperative (\emph{S.C.P.P. Buzos
y Pescadores de la Baja California}) that promotes co-management of
marine resources {[}49,50{]}. Additionally, fishers have Territorial Use
Rights for Fisheries (TURFs) that provide them with exclusive access
rights to exploit the benthic marine resources within a given perimeter
{[}50{]}.

\textbf{Fig 3. General location of Isla Natividad (left) and map of the
island (right).} The marine reserve polygon is indicated in red, and the
approximate location of control sites is indicated by blue squares (B =
Babencho, D = La Dulce).

In 2006, the Isla Natividad community implemented two community--based
marine reserves within their TURF {[}9,51,52{]}. Evidence suggest that
these reserves have been effective at enhancing resilience to climate
variations {[}9{]} and preserving genetic diversity of high value
commercial species such as abalone {[}11{]}. These ecological benefits
have been translated into economic benefits, enhancing population
persistence and bolstering abalone fisheries {[}53{]}. For the purpose
of this evaluation, we focused on the ``La Plana / Las Cuevas'' marine
reserve, located at the southern end of the island (Fig. 3) and its
corresponding control site ``La Dulce / Babencho''.

@@

Local fishers (who were trained in scientific diving by the CSO
Comunidad y Biodiversidad, A.C. (COBI; \url{www.cobi.org}), ReefCheck
California, and Stanford University) and personnel from these
institutions performed SCUBA dives to record fish and invertebrate
richness and abundances, as well as fish total length. They recorded
information along 30 m transects, with a sampling window of 2 m x 2 m
following a standardized ReefCheck protocol {[}54{]}. Ecological surveys
were performed yearly in each reserve and corresponding control site(s),
before and after the implementation of the reserve, providing the
requisite time series data inside the reserve and for a suitable control
site. Annual surveys (2006--2016) were carried out in late July -- early
August, performing a total of 242 and 245 transects in the reserve site
for fish and invertebrate surveys, respectively. Similar sampling effort
was applied to the control site, with 221 fish and 222 invertebrate
transects. Between 12 and 27 transects were performed in each site every
year.

>>>>

Empirical evidence shows that MPAs increase biomass {[}4,7{]}, enhance
resilience to climatic impacts {[}8,9{]}, and preserve genetic diversity
{[}10{]}. Compared to MPAs that grant partial protection, marine
reserves have higher levels of biomass, density, richness, and larger
organisms {[}3,11--13{]}. However, these effects are often measured as
biological changes within the reserves through time, and many lack a
control site for comparison {[}14{]}. This approach does not account for
other factors (\emph{e.g.} system-level changes in productivity ;
{[}15{]}) for which one must control {[}16{]} in order to causally
attribute a biological change to the reserve. Other studies have used a
control--impact comparison approach that uses control sites but does not
address temporal variability {[}4,7,17--19{]}. A smaller fraction of
studies have used a before--after--control--impact (\emph{i.e.} BACI)
design comparing reserves to control sites before and after
implementation {[}4,20,21{]}, which allows the use of causal inference
techniques that estimate the effect of the reserve. But even when proper
causal inference can be drawn, results are often different across
reserves. Effects of reserves on ecological and economic outcomes are
highly heterogeneous, and often depend on the specific ecological,
economic, and social context.

Standardization of marine reserve evaluation is not new. The IUCN
framework ``How is your MPA doing?'' {[}22,23{]} provides a
comprehensive list of biological, socioeconomic, and governance
indicators, and insights into how these may be measured or collected.
But this framework stops short of analysis, and provides a user with
little guidance about establishing causal inference about the reserve.
Recent work {[}24{]} integrates these three dimensions via the Social
Ecological Systems Framework {[}25,26{]} and suggests the use of causal
inference techniques to provide a measure of the effect of conservation
interventions. However, neither of these approaches provide a
user-friendly tool that ensures replicability and scalability of the
analysis, particularly when used by the fishers and decision makers
themselves.

@@

The purpose of this paper is to describe a user--friendly tool, called
``MAREA'', to rigorously systematize the evaluation of marine reserve
effectiveness in terms of fisheries and marine conservation. The tool is
in the form of an open--source application that uses state--of--the--art
methods from program evaluation to compare a reserve to control sites
along a number of ecological, economic, and governance dimensions. We
first provide a list of commonly stated management objectives and match
them to appropriate indicators. We then develop a simple approach to
analyzing these indicators building on causal inference techniques
{[}20{]}, which help us understand the effect of management
interventions {[}24,37{]}. To implement the analytical approach, we
introduce the Marine Reserve Evaluation Application (MAREA), an open
source, web--based tool that automates the framework described in this
paper and enables its broader use. Finally, we present a case study on
the evaluation of a marine reserve established by the fishers of Isla
Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.

@@

Throughout this study, we will refer to the stated goals for which a
marine reserve was designed as ``objectives.'' This work was motivated
by the case of Mexico, where 39 reserves have been implemented over the
past five years to achieve objectives such as increasing productivity in
nearby waters or recovery of overexploited species; most of these
reserves have never been formally evaluated for effectiveness at meeting
those objectives. Thus, our focus was on identifying common objectives
of marine reserves in Mexico. However, a literature review and
discussions with marine reserve researchers suggest that the objectives
driving Mexican marine reserve implementation are similar to those in
the rest of the world. Thus, we group these objectives into seven major
categories, which may be applied to marine reserves worldwide. Any given
reserve may have been implemented to meet one or more of these. The list
includes objectives stated in legislation {[}38,39{]} and official
documents such as the Technical Justification Studies (\emph{Estudios
Técnicos Justificativos}), agreements, and decrees associated to these
areas:

@@

Based on these seven objectives, we determined a set of associated
indicators to evaluate reserve effectiveness. These indicators are
specific variables on which data could be collected, and analyzed, to
ultimately determine whether the corresponding objective was causally
being achieved by the marine reserve. The list of indicators was
compiled through a review of scientific literature in which we
identified indicators that were used to measure similar
objectives{[}3--5,7,11,13,14,17--20,22,23,40--43{]}. A first filter
eliminated indicators for which baseline data do not typically exist in
Mexico. The preliminary list of indicators was reviewed at a workshop
with participation of members from Mexican fishery management agencies
and non-government organizations. Later, these were presented to fishers
from the Ensenada Fishing Cooperative (\emph{S.C.P.P. Ensenada}), in El
Rosario, Baja California, who provided input. Our final list of
indicators includes those identified in review works {[}4,43{]}.

Indicators are divided into three main categories: biological,
socioeconomic, and governance (Table 1). The nine biological indicators
focus on fish and invertebrate communities that are evaluated using
underwater ecological surveys performed inside and outside the reserve
(see Data and Analysis section for specific sampling design and
methodologies). Five socioeconomic indicators reflect the performance of
the fishery in terms of landings, income from landings, and availability
of alternative livelihoods. Fifteen governance indicators describe the
governance structures under which the community operates (\emph{e.g.},
access rights to the fishery, number of fishers, legal recognition of
the reserve). Some indicators are quantitative and require a numerical
entry (\emph{e.g.} Fish biomass) while others are qualitative and rely
on a descriptive entry (\emph{e.g.} Reasoning for reserve location).
Many of them specifically measure an outcome of the reserve, though some
are designed to further the understanding of the mechanisms driving a
reserve's performance. In that sense, most biological and socioeconomic
indicators are outcome variables. On the other hand, governance
indicators are viewed as possible explanatory variables of reserve
performance. Whenever an indicator is applied to ``Target species'', it
means that the indicator can be used for all species (\emph{e.g.} Fish
Biomass) and/or for individual species that are either the conservation
target of the reserve or are of particular economic or ecological
interest (\emph{e.g.} Grouper Biomass). Finally, indicators B3 and B4
are different in that B3 look only at the density of organisms above
size at first maturity (related to reproductive potential), while B4
measures the density of all fish or of a target species. Each indicator
targets different plausible desired outcomes, like increased
reproductive potential (\emph{i.e.} B3; {[}44{]}) or having more fish
-regardless of their size- to atract tourism (\emph{i.e.} B4). Table 1
presents the proposed indicators, and Table 2 shows how objectives are
matched with biological and socioeconomic indicators. Governance
indicators are excluded from Table 2, as we suggest that all of them
should be used in an evaluation, regardless of the objectives.

@@

\caption{\label{tab:unnamed-chunk-2}List of indicators to evaluate the effectiveness of no-take marine reserves.}
\centering
\resizebox{\linewidth}{!}{\begin{tabular}[t]{l|l|l|l}
\hline
Code & Indicator & Data type & Unit\\
\hline
\multicolumn{4}{l}{\textbf{Biological}}\\
\hline
\hspace{1em}B1 & Shannon diversity index & Continuous & \\
\hline
\hspace{1em}B2 & Species richness & Discrete & Number of species/transect\\
\hline
\hspace{1em}B3 & Density of mature organisms & Continuous & Percent\\
\hline
\hspace{1em}B4 & Density* & Continuous & Organisms/transect\\
\hline
\hspace{1em}B5 & Natural Disturbance & Descriptive & \\
\hline
\hspace{1em}B6 & Mean Trophic Level & Continuous & \\
\hline
\hspace{1em}B7 & Biomass* & Continuous & kg/transect\\
\hline
\multicolumn{4}{l}{\textbf{Socioeconomic}}\\
\hline
\hspace{1em}S1 & Total landings* & Continuous & kg\\
\hline
\hspace{1em}S2 & Income from total landings* & Continuous & \$\\
\hline
\hspace{1em}S3 & Alternative economic opportunities & Ordinal & \\
\hline
\multicolumn{4}{l}{\textbf{Governance}}\\
\hline
\hspace{1em}G1 & Access to the fishery & Categorical & \\
\hline
\hspace{1em}G2 & Number of fishers & Discrete & \\
\hline
\hspace{1em}G3 & Legal recognition of reserve & Binary & \\
\hline
\hspace{1em}G4 & Reserve type & Descriptive & \\
\hline
\hspace{1em}G5 & Illegal harvesting & Ordinal & \\
\hline
\hspace{1em}G6 & Management plan & Binary & \\
\hline
\hspace{1em}G7 & Reserve enforcement & Descriptive & \\
\hline
\hspace{1em}G8 & Size of reserve & Discrete & \\
\hline
\hspace{1em}G9 & Reasoning for reserve location & Descriptive & \\
\hline
\hspace{1em}G10 & Membership to fisher organizations & Binary & \\
\hline
\hspace{1em}G11 & Type of fisheries organizations & Categorical & \\
\hline
\hspace{1em}G12 & Representation & Ordinal & \\
\hline
\hspace{1em}G13 & Internal Regulation & Binary & \\
\hline
\hspace{1em}G14 & Perceived Effectiveness & Categorical & \\
\hline
\hspace{1em}G15 & Social Impact of Reserve & Categorical & \\
\hline
\multicolumn{4}{l}{\textsuperscript{*} The indicator is also applied to target species}\\
\end{tabular}}
\end{table}

@@

\caption{\label{tab:unnamed-chunk-3}Management objectives and respective performance indicators.}
\centering
\resizebox{\linewidth}{!}{\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
\hline
Objective & B1 & B2 & B3 & B4 & B4* & B5 & B6 & B7 & B7* & S1 & S1* & S2 & S2* & S3\\
\hline
Avoid overexploitation &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
\hline
Conserve species under a special protection &  &  & x &  & x & x &  &  & x & x &  & x &  & x\\
\hline
Maintain biological process & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
\hline
Improve fishery production in nearby waters &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
\hline
Preserve biological diversity and the ecosystem & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
\hline
Recover overexploited species &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
\hline
Recover species of economic interest &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
\hline
\multicolumn{15}{l}{\textsuperscript{*} The indicator is applied to target species}\\
\end{tabular}}
\end{table}

@@

In many coastal marine reserves, biological data are often collected via
underwater visual censuses as part of a reserve's monitoring program.
Scientific divers record fish and invertebrate richness and abundances,
as well as fish total length along belt transects. Ecological surveys
are typically performed annually in each reserve and corresponding
control site(s), before and after the implementation of the reserve,
providing a sampling design that can be used to draw causal inference.
Control sites are areas where habitat is similar to that of the reserve,
but with presence of fishing activity; in principle these are areas that
are otherwise observationally identical to the reserve site, but where,
for presumably random reasons, a reserve was not implemented. While
transect dimensions (\emph{i.e.} length and width) and sampling methods
might vary from study to study, the general idea remains the same:
richness, abundances, and sizes of organisms are recorded in a
study--specific standardized way. For this reason, MAREA does not assume
specific transect dimensions, and pertinent indicators are calculated
per transect (Table 1). More information on data collection and
formatting is provided in a guidebook {[}45{]}, which is available in
English and Spanish in MAREA's interface.

This sampling design for biological data allows us to use causal
inference techniques {[}20,46{]} to evaluate the effect of the reserve
on biological indicators. The hypothesis that the indicators will
respond to implementation of the reserve is tested by analyzing spatial
and temporal changes in each numeric biological indicator (all but B5)
using generalized linear models {[}20{]}. To account for variations in
the environment and survey conditions, covariates that are gathered
during the underwater ecological surveys are included in the
difference-in-differences model with form:

@@

In this model, \(i\), \(t\), and \(z\) are indices for transect, time,
and zone (control or reserve site), respectively. This model allows us
to estimate the change in an indicator (\(I\)) based on the year
(\(Y\)), a dummy variable that indicates treatment (\(Z\); \emph{i.e.}
control or reserve), an interaction between a dummy variable that
indicates before or after implementation (\(P\)) and treatment (\(Z\)),
and covariates such as bottom temperature (\(T\); in °C), horizontal
visibility during the survey (\(V\); in m), and depth at which survey
was performed (\(D\); in m). \(\epsilon\) represents the error term
associated to the regression. Here, years are modeled as factors, using
the first year as the reference level. This does not impose a linear
structure in the way an indicator changes through time (\emph{i.e.} the
change in biomass between 2006 and 2007 does not have to be the same as
the change between 2015 and 2016). The treatment and implementation
variables, modeled as dummy variables, are coded as Control = 0 and
Reserve = 1; and Before implementation = 0 and After implementation = 1,
respectively.

Socioeconomic data are often collected by fishers, natural resource
management agencies, or CSOs by recording landings, income, and
sometimes prices for each species. To control for inflation, income is
adjusted with the country's consumer price index {[}47{]}:

@@

While this model does not formally allow for causal inference, we can
still measure changes in mean landings and income before and after the
implementation of the reserve and provide valuable input. For both
models (eq. 1 and eq. 3), we estimate the model coefficient with
ordinary least squares, and calculate heteroskedastic--robust standard
errors {[}48{]}.

While biological and some economic data are regularly collected,
governance data are typically not available nor systematically collected
by the community or other organizations. Therefore, we created a survey
specifically designed to collect information needed for the proposed
indicators (B5, S3, and G1--G15). The survey is included as
supplementary material in English (S1 Appendix) and Spanish (S2
Appendix). To analyze governance information, we developed a framework
based on a literature review of common governance structures and their
relation to effectiveness in managing fisheries or marine reserves (S3
Table). This approach has been proven to successfully evaluate
governance structures {[}49{]}. Unlike with biological and socioeconomic
objectives (see eqs 1 and 3), MAREA does not quantitatively analyze
governance information. Rather, it is presented along with the
biological and socioeconomic indicators to provide managers and users
with a more complete description of the reserve.

@@

We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}50{]} using
the Shiny package {[}51{]}, to build an interactive web application
hosted on an open server; the MAREA app can be accessed at
\href{turfeffect.shinyapps.io/marea/}{turfeffect.shinyapps.io/marea}.
While the original version was developed in Spanish because it was aimed
for Mexico and other Latin-American countries, all of its content can be
translated by a translation widget available within the app.

MAREA is designed as a 6--step process, divided in tabs that appear upon
launching the app. The first tab introduces the app and summarizes the
evaluation process. Then, the user selects management objectives, which
MAREA automatically matches to appropriate indicators, based on Table 2.
Users can also select and un-select indicators based on their interests
and data availability by ``clicking'' on the check-boxes in MAREA. The
user can then load data on one or more reserves, using standard *.csv
text files; sample datasets are provided within MAREA. Once data have
been loaded, MAREA identifies all reserves in the data, and lets the
user select the reserve to be evaluated. At this point, the user can
also specify the year of implementation of the reserve, reserve
dimensions, and indicate target species that are of particular
management interest. MAREA provides the user with a section to confirm
that all the decisions made leading up to that point are correct. Once
the user has confirmed all input data, objectives, and other
information, MAREA performs the formal program evaluation analyses
discussed above. For a typical data set, the automated analysis step
takes less than one second. Finally, the user is taken to the results
tab where all results are presented in a simple format. The user can
also download a more comprehensive technical report produced in *.pdf
format.

@@

The scorecard is produced with functions from the Shinydashboard package
{[}52{]}. The technical report is produced by a parameterized Rmarkdown
document {[}53{]} processed by the knitr package {[}54{]}. Another
feature of MAREA is that the user can choose to share the data. Once the
technical report is downloaded, the information on the reserve, its
management objectives, and all uploaded data are saved into a central
repository. These data can be accessed at any time by any person
interested in acquiring them at \url{github.com/turfeffect/MAREAdata}.

@@

While MAREA is a general tool that can be easily employed to evaluate
the effectiveness of any marine reserve with the required input data, we
illustrate its use here by applying it to one marine reserve near Isla
Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8
Km off the Pacific Coast of the Baja California Peninsula (Fig. 3),
where fishers operate under a fishing cooperative (\emph{S.C.P.P. Buzos
y Pescadores de la Baja California}) that promotes co-management of
marine resources {[}55,56{]}. Additionally, fishers have Territorial Use
Rights for Fisheries (TURFs) that provide them with exclusive access
rights to exploit the benthic marine resources within a given perimeter
{[}56{]}.

\textbf{Fig 3. General location of Isla Natividad (left) and map of the
island (right).} The marine reserve polygon is indicated in red, and the
approximate location of control sites is indicated by blue squares (B =
Babencho, D = La Dulce). Shapefiles for Mexican coastline and the US
were obtained from the Instituto Nacional de Estadistica, Geografia e
Informatica of Mexico (INEGI:
www.inegi.org.mx/geo/contenidos/geoestadistica/m\_geoestadistico.aspx)
and the tmap R package {[}57{]}, respectively.

In 2006, the Isla Natividad community implemented two community--based
marine reserves within their TURF {[}8,40,58{]} after establishing a
baseline for the soon--to--be reserves and control sites. Evidence
suggest that these reserves have been effective at enhancing resilience
to climate variations {[}8{]} and preserving genetic diversity of high
value commercial species such as abalone {[}10{]}. These ecological
benefits have been translated into economic benefits, enhancing
population persistence and bolstering abalone fisheries {[}42{]}. For
the purpose of this evaluation, we focused on the ``La Plana / Las
Cuevas'' marine reserve, located at the southern end of the island (Fig.
3) and its corresponding control site ``La Dulce / Babencho''.

@@

Local fishers (who were trained in scientific diving by the CSO
Comunidad y Biodiversidad, A.C. (COBI; \url{www.cobi.org}), ReefCheck
California, and Stanford University) and personnel from these
institutions performed SCUBA dives to record fish and invertebrate
richness and abundances, as well as fish total length. They recorded
information along 30 m transects, with a sampling window of 2 m x 2 m
following a standardized ReefCheck protocol {[}59{]}. Ecological surveys
were performed yearly in each reserve and corresponding control site(s),
before and after the implementation of the reserve, providing the
requisite time series data inside the reserve and for a suitable control
site. Annual surveys (2006--2016) were carried out in late July -- early
August, performing a total of 242 and 245 transects in the reserve site
for fish and invertebrate surveys, respectively. Similar sampling effort
was applied to the control site, with 221 fish and 222 invertebrate
transects. Between 12 and 27 transects were performed in each site every
year.
